Dynamics of Laterally Propagating Flames in X-ray Bursts

### 2019-10-26 xrb/paper.tex
Added some comments and rephrased a few sentences

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
We investigate the structure of laterally-propagating flames through
the highly-stratified burning layer in an X-ray burst.  Simple \MarginPar{is simple the right word here?}
two-dimensional simulations of flame propagation are performed through
a rotating plane-parallel atmosphere, exploring the structure of the
flame.  We discuss the approximations needed to capture the length and
time scales at play in an X-ray burst.  Our studies complement the
other multidimensional studies of burning in X-ray bursts.
\end{abstract}

@@

X-ray bursts (XRBs) are thermonuclear explosions of an accreted H or
He layer on the surface of a neutron star (see~\citealt{galloway:2017}
for a review).  Observations of bursts can help constrain the
properties of the underlying neutron star, helping to illuminate the
nuclear equation of state~\citep{steiner:2010,ozel2016masses}.  Extensive
observations of brightness oscillations during the rise have provided
evidence that the burning begins in a localized region and spreads
over the surface of the neutron
star~\citep{bhattacharyya:2006,bhattacharyya:2007,chakraborty:2014}.

One dimensional studies of XRBs are very successful in predicting the
lightcurves and recurrance times (see, e.g., \citealt{woosley-xrb}).
These assume spherical symmetry and thus cannot capture the effects of
localized burning spreading across the neutron star.  These studies
have also been used to explore the sensitivity of the burst
observables to accretion and reaction
rates~\citep{cyburt:2010,Jose2010a,Lampe2016}, and model individual
bursts~\citep{johnston:2019}.

Multidimensional simulations of burning on a neutron star are more
difficult, with both the temporal and spatial scales presenting
challenges (see \citealt{astronum_2018} for an overview).  For the spatial scales, we need to
resolve the reaction zone, $\mathcal{O}(10~\mbox{cm})$ or smaller, the
scale height of the atmosphere, $\mathcal{O}(500~\mbox{cm})$, and the
Rossby scale where the Coriolis force balances the lateral pressure
gradient, $\mathcal{O}(10^5~\mbox{cm})$~\citep{spitkovsky2002}.  For
the temporal scales, capturing the rise, $\mathcal{O}(1~\mbox{s})$,
and the decay of the burst, $\mathcal{O}(10~\mbox{s})$, and the
accretion period between bursts, $\mathcal{O}(10^4~\mbox{s})$ is
beyond the ability of multidimensional hydro codes currently.
Nevertheless, significant progress has been made in understanding the
multidimensional nature of XRBs, through various approximations.x

Laterally propagating detonations were modeled by
\citet{fryxellwoosley82,hedet}, but these would only be applicable at
very high densities, since it is difficult to detonate helium at the
densities found in normal XRBs, and even harder to detonate hydrogen
because of the waiting times for weak reactions.  This means that
detonations may only really be applicable to superbursts (where carbon
is the reactant)~\citep{Weinberg2006b,Weinberg2007}.

Global multidimensional studies were performed by
\citet{spitkovsky2002}, where it was demonstrated that the Coriolis
force plays an important role in confining the burning as it spreads
across the neutron star surface.  These calculations used the
shallow-water approximation, so the vertical details of the
atmosphere's structure we not modeled.  The model showed that the
horizontal pressure gradient between the ash and fuel can be important
in accelerating the burning front.

Small-domain studies of convective burning in XRBs preceding flame
development have been done in two-dimensions \citep{lin:2006,xrb,xrb2}
and three-dimensions \citep{xrb3d}.  These calculations used low Mach
number methods that approximate the hydrodynamics equations to filter
soundwaves, enabling large timesteps and efficient modeling of
subsonic convection.  While these calculations could not support
the lateral differences needed for flame spreading, they can help
understand the role that convection plays in distributing the initial
burning products vertically throughout the neutron star atmosphere as well
as the nature of any turbulence the burning front might encounter.

@@

For deflagrations, we either need to resolve the structure of the
reaction zone or use a flame model.  Flame models usually assume that
the flame structure is thin compared to the size of the system (see,
e.g., \citet{Ropke2007} for applications to Type Ia supernovae).  For
XRBs however, the flame thickness is comparable to the scale height of
the atmosphere, so we cannot use these approximations.  Accurate
models of flames therefore require that we resolve the thermal width,
which is $\mathcal{O}(10~\mbox{cm})$ for helium flames~\citep{Timmes00}.

The goal of this study is to understand what numerical and physical
approximations are needed to perform a full hydrodynamical,
multidimensional simulation of flame propagation through the
atmosphere on a neutron star.  For simplicity in this first set of
calculations, we will use pure helium composition.  These studies
complement the other multidimensional work described above in helping
us to build a picture of the dynamics of X-ray bursts.

@@

All simulations are performed with the \castro\ hydrodynamics
code~(\citealt{castro}; see also \citealt{astronum:2017} for a recent
description).  We evolve the system of fully compressible Euler
equations for reacting flow:
\begin{eqnarray}
\frac{\partial( \rho X_k)}{\partial t} &=& - \nabla\cdot (\rho \Ub X_k) + \rho \omegadot_k \\
%
\frac{\partial (\rho \Ub)}{\partial t} &=& - \nabla\cdot (\rho \Ub \Ub) - \nabla p +
    \rho \gb \nonumber \\
  &&-2 \rho \Omegab\times \Ub - \Omegab \times (\Omegab \times \rb) \\
%
\frac{\partial (\rho E)}{\partial t} &=& - \nabla \cdot (\rho \Ub E + p\Ub) +
    \nabla \cdot \kth \nabla T + \rho \epsdot +\nonumber\\
  && \rho \Ub \cdot \gb - \rho (\Omegab \cdot \rb)(\Omegab \cdot \Ub) + \rho |\Omegab|^2 (\Ub \cdot \rb)
\end{eqnarray}
Here, $\rho$ is the mass density, $\Ub$ is the velocity, and $E$ is
the specific total energy, which is related to the specific internal
energy as $e = E - |\Ub|^2/2$.  Species are described by mass
fractions, $X_k$ (such that $\sum_k X_k = 1$), and creation rates,
$\omegadot$, and are related to the total specific energy generation
rate, $\epsdot$.  The total mass conservation,
\begin{equation}
\frac{\partial \rho}{\partial t} = - \nabla\cdot (\rho \Ub)
\end{equation}
implies $\sum_k \omegadot_k = 0$.  \castro\ uses an unsplit piecewise
parabolic method (PPM) with characteristic tracing for solving the
hydrodynamics~\citep{ppm,millercolella:2002}, generalized to an
arbitrary equation of state~\citep{zingalekatz}.  Reactions are
incorporated via Strang splitting~\citep{strang:1968}, giving a method
that is overall second-order accurate in space and time.
\castro\ uses the \amrex\ adaptive mesh refinement
library~\citep{amrex_joss} to manage a hierarchy of grids at different
resolutions.

An equation of state of the form $p = p(\rho, e, X_k)$ completes the
thermodynamic description of the system.  Finally, $\kth$ is the
thermal conductivity and $T$ is the temperature.  Diffusion is treated
explicitly in time.

@@

We use a general stellar equation of state with nuclei (treated as an
ideal gas), photons, and degenerate/relativistic electrons, as
described in \cite{timmes_swesty:2000}.  For reactions, we use a 13
isotope alpha chain, derived from the {\tt aprox13}
network~\citep{timmes_aprox13}.  For one of our runs, we use the
smaller 7-isotope network described in \citet{iso7}.  We integrate the
network using the VODE integration package~\citep{vode}, and our
implementation is provided in the StarKiller Microphysics
source~\citep{starkiller}.
We note that we do not explicitly model viscosity.  The reactions will
provide the smallscale cutoff to the instabilities and turbulence at
the flame front.  We also do not include species
diffusion---astrophysical flames tend to have large Lewis numbers, so
this is not expected to be important~\citep{timmeswoosley:1992}.
Finally, we use the thermal conductivities described in
\citet{Timmes00}.

All simulations use adaptive mesh refinement to refine on the
atmosphere (leaving the space between the top of the atmosphere and
upper boundary at low resolution).  In addition to the base grid, we
use up to 3 refinement levels, the first one a factor of 4 finer than
the previous and the remaining each a factor of 2 finer than the
previous.  \castro\ subcycles in time, so the finer grids are evolved
with a finer timestep than the coarse grids.  Occasionally the
timestep chosen at the start of a cycle will violate the CFL condition
during the advance of the finer grids.  In this case, we restart the
finer grid evolution with a smaller timestep, subcycling within the
larger timestep hierarchy.  We use a CFL number of 0.8 for our
simulations.  
% \MarginPar{true for all?} 
The base grid for our standard
simulations is $768\times 192$ zones and the corresponding finest grid
when 3 refinement levels are added would be $12288\times 3072$ zones.
Our standard domain is $1.2288\times 10^5~\mathrm{cm} \times
3.072\times 10^4~\mathrm{cm}$, so we have $10~\mathrm{cm}$ resolution
on the finest grid.  We only refine the fuel layer in the left half of
the domain at the highest resolution (and only down to densities of
$2.5\times 10^4~\gcc$), since this is where we expect the flame to
propagate.  At the start of the simulation, $3.4\%$ of the domain is
at the finest resolution.  This increases to $7.4\%$ by the end of the
simulation, because of the increase in the scale height of the atmosphere
behind the flame.  \MarginPar{add a pic of the grid to show refinement}

Thermal diffusion is modeled explicitly, using a predictor-corrector
scheme to achieve second-order accuracy.  A verification test of the
diffusion scheme is shown in Appendix~\ref{app:diffusion}.  The
explicit thermal diffusion requires a timestep limiter of the form:
\begin{equation}
\Delta t_\mathrm{diff} \le \frac{1}{2} \frac{\Delta x^2}{\mathcal{D}}
\end{equation}
where $\mathcal{D} = \kth/(\rho c_v)$ is the thermal diffusivity.  The
diffusivity increases rapidly at the top of the atmosphere, making
these low density regions determine the overall timestep for the
simulations.  Therefore, we disable thermal conduction at low
densities where it is not expected to be important.

We use hydrostatic boundary conditions on the lower boundary, using
a discretized hydrostatic equilibrium equation of the form:
\begin{equation}
\label{eq:hse}
p_i = p_{i-1} + \frac{1}{2} \Delta y (\rho_i + \rho_{i-1}) g
\end{equation}
holding the temperature constant in the ghost cells.  This is solved
together with the equation of state.  The velocity is reflected here.
This procedure follows the form described in \citet{ppm-hse}.  The
left boundary is reflecting and the right boundary is a zero-gradient
outflow.  The top boundary sets the state to simply the conditions
in our outer buffer region of the initial model (see below) with the
normal velocity set to the larger of zero or the velocity at the top
of the domain (this prevents incoming velocities at the top) and the
transverse velocities set to zero.

When we begin the simulation, there is a transient phase as the flame
gets established.  Material that is forced upward will encounter the
steep density gradient at the top of the atmosphere and accelerate as
it is blown out of the atmosphere.  Eventually this material will fall
back to the top of the atmosphere.  In this paper, we are mostly
concerned with the behavior of the flame and not any material that is
violently blown out of the top of the atmosphere, so we apply a sponge
to this region.  This is similar to what we've done in \cite{xrb3d}.
The sponge drives the velocity of the material in the low
density \MarginPar{add mathematical form of sponge} regions at the top
of the atmosphere to zero.  This sponging helps increase our timestep
as well.

@@

The speed and thickness of a laminar helium flame are determined by the
energy generation rate and conductivity, and scale roughly as 
\begin{equation}
v_f \approx \sqrt{k_\mathrm{th} \dot{\epsilon}} \qquad
\lambda_f \approx \sqrt{\frac{k_\mathrm{th}}{\dot{\epsilon}}}
\end{equation}
\citep{orourke:1979,khokhlov:1993}.
At the densities we consider in this simulation, the pure He flame
speed is quite slow and would require a prohibitively long integration
time to see significant evolution of the burning.  We therefore
consider boosted flames in this first paper, to accelerate the burning
and allow us to understand the qualitative effects of laterally
propagating flames.  To boost the flame, while keeping the thickness
the same, we can multiply both the burning rate and the conductivity
by the same factor.  For our standard calculations, we choose 10 for
each, to give a flame 10$\times$ faster.  We call this the ``10/10''
flame.  We will also do a simulation with the reactions and
conductivity both boosted by 5, the ``5/5'' flame.

@@

The mass fractions use this same profile, switching from a set
describing the underlying star, ${X_k}_\star$, and the set for the
accreted material, ${X_k}_\mathrm{atm}$, which is used in the
isentropic and outer regions.  Note, since the profile above is
linear in $X$, if the initial mass fractions sum to one, then the blended
mass fractions in the ramp region also sum to one.

@@

Creating the model involves specifying $T_\star$, $T_\mathrm{hi}$,
$T_\mathrm{lo}$, $\rho_\mathrm{int}$, $H_\star$,
$\delta_\mathrm{atm}$, ${X_k}_\star$, and ${X_k}_\mathrm{atm}$.  We
then integrate outwards from the base of the ramp region ($y =
H_\star$), enforcing the discrete form of hydrostatic equilibrium,
Eq.~\ref{eq:hse}.  Integrating upwards, we would find $p_i$ and
$\rho_i$ using a Newton solve together either with the temperature
specified, $T_i = T(p_i, \rho_i, \{X_k\}_i)$ (in the isothermal, ramp,
and buffer layers) or constant entropy, $s_i = s(p_i, \rho_i,
\{X_k\}_i)$, in the fuel layer.  In each case, using the equation of
state.  This follows the procedures described in \citet{ppm-hse}.  We
use the constant temperature for all $y < H_\star +
3\delta_\mathrm{atm}$.  Above this, we switch to isentropic until the
temperature drops to a floor value, $T_\mathrm{lo}$, at which point we
again keep the temperature constant.  The integration of the
atmosphere continues until the density falls to a low density cutoff,
$\rho_\mathrm{cutoff}$.  The material above this height is taken to
have constant density and temperature.

@@

We create two models, a ``cool'' model representing the atmosphere
ahead of the flame and a ``hot'' (or perturbed) model that will
initiate the flame by driving the burning.  The parameters we use for
the model generation are listed in Table~\ref{table:params}. \MarginPar{reference and discuss the figure of initial model here}

@@

Table~\ref{table:sim_names} summarizes the simulations that we ran.
The majority of them used a reaction rate boosting of $10$ and a
conductivity boosting of $10$, which should increase the flame speed
by a factor of $10$.  Most simulations used a resolution of
$10~\mbox{cm}$ and a domain width of a little more than one kilometer.  
We use an artifically high rotation rate of
$2000~\mbox{Hz}$, which gives a Rossby length of
\begin{equation}
L_R = \frac{\sqrt{g H_0}}{\Omega} \sim 3\times 10^4~\mathrm{cm}
\end{equation}
using a scale height $H_0 = 10^3~\mathrm{cm}$.  This is about one
quarter of domain width.  The run at $20~\mbox{cm}$ resolution used
one fewer level of refinement.  The slower rotating case using a
slightly wider domain to accommodate the expected larger Rossby
length.  We note that the entire simulation framework for these
calculations is freely available in the \castro\ github
repository\footnote{\url{https://github.com/AMReX-Astro/Castro}, using
  the {\tt flame\_wave} setup.}.  In the discussions below, we'll use
the simulation names defined in the table to refer to specific runs
and we'll use the 10/10 run as the reference calculation.

@@

Figure~\ref{fig:time_series} shows the time evolution of the 10/10
simulation, focusing on the mean-molecular weight,
\begin{equation}
\bar{A} = \left ( \sum_k \frac{X_k}{A_k} \right )^{-1}
\end{equation}
In each frame the buffer of \isot{Ni}{56} that serves as the
underlying neutron star is seen spanning the bottom of the domain.
Above that, the composition begins as pure \isot{He}{4}, but as the
simulation progresses, the flame processes this to heavier nuclei,
increasing $\bar{A}$.  By about $10~\mbox{ms}$, we see the flame front
is reasonable well-defined.  We see that the bottom of the burning
front is lifted off of the base of the atmosphere, greatly increasing
the surface area of the burning compared with a perfectly vertical
flame front.  By $20~\mbox{ms}$ the flame has moved out substantially
and we are beginning to see a gradient in the composition of the ash,
with the heavier nuclei furthest behind the flame.  The boosting of
the burning likely artificially increases this effect. \MarginPar{Doesn't the increased flame speed of the 10/10 run vs. the 5/5 run show this?}

@@

A final feature worth noting is the ash that seems to move along the
surface at a higher velocity than the flame, via surface gravity
waves.  The sponging that we perform is likely damping this to some extent, and
the method by which we initialize the flame may induce a larger
transient than in nature, nevertheless, this surface ash is intriguing
because it affects the composition of the photosphere ahead of the
burning front, potentially changing our interpretation of
observations.  This is something that will be explored more fully in
the future.

@@

Our default resolution puts $\sim 80$ zones vertically in the
``cool'' \MarginPar{double check} model atmosphere height.  To
understand the effects of resolution, we also performed a run with one
fewer level, giving a $20~\mbox{cm}$ resolution overall (this is our
10/10-lores run).  Figure~\ref{fig:10_10_lowres} shows the fields at
20~ms.  The structure is largely the same as the 10/10 run, with
largely the same flame shape and position, and the same structure in
the energy generation rate.  Since we do not have a jump in refinement
right below the atmosphere, we don't see the temperature feature from
the initial model mapping there, but we do see some cooling in the
\isot{Ni}{56} region.  This is likely a resolution effect, again due
to the strong gradient in temperature at the base of the atmosphere.
The strong agreement in the low resolution run to the 10/10 run gives
us confidence that we are capturing the flame physics properly.

@@

We can also look at the effect of the network size.
Figure~\ref{fig:network} shows a comparison of the 10/10 boosting run
with the standard 13-isotope ``aprox13'' network and the reduced
7-isotope ``iso7'' network.  We see that the flame in the aprox13 case
is slightly more advanced and has a higher $\bar{A}$ than the iso7
case.  We see in the next section that these two networks give largely
the same flame speed.  The reduced network size saves a lot of memory
which will be important when we transition to 3-d simulations.

>>>>

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
We investigate the structure of laterally-propagating flames through
the highly-stratified burning layer in an X-ray burst.  Simple \MarginPar{is simple the right word here?} \MarginPar{It might be better to say simulations are performed through a simplified atmosphere?}
two-dimensional simulations of flame propagation are performed through
a rotating plane-parallel atmosphere, exploring the structure of the
flame.  We discuss the approximations needed to capture the length and
time scales at play in an X-ray burst.  Our studies complement 
other multidimensional studies of burning in X-ray bursts.
\end{abstract}

@@

X-ray bursts (XRBs) are thermonuclear explosions in an accreted H or
He layer on the surface of a neutron star (see~\citealt{galloway:2017}
for a review).  Observations of bursts can assist in constraining the
properties of the underlying neutron star, helping to illuminate the
nuclear equation of state~\citep{steiner:2010,ozel2016masses}.  Extensive
observations of brightness oscillations during the rise (the initial phase of the burst where the observed flux rapidly increases) have provided
evidence that the burning begins in a localized region and spreads
over the surface of the neutron
star~\citep{bhattacharyya:2006,bhattacharyya:2007,chakraborty:2014}.

One dimensional studies of XRBs have been very successful in predicting the
lightcurves and recurrence times (see, e.g., \citealt{woosley-xrb}).
These assume spherical symmetry and thus cannot capture the effects of
localized burning spreading across the neutron star.  These studies
have also been used to explore the sensitivity of the burst
observables to accretion and reaction
rates~\citep{cyburt:2010,Jose2010a,Lampe2016}, and to model individual
bursts~\citep{johnston:2019}.

Multidimensional simulations of burning on a neutron star are more
difficult, with both the temporal and spatial scales presenting
challenges (see \citealt{astronum_2018} for an overview).  For the spatial scales, we need to
resolve the reaction zone, $\mathcal{O}(10~\mbox{cm})$ or smaller, the
scale height of the atmosphere, $\mathcal{O}(500~\mbox{cm})$, and the
Rossby scale where the Coriolis force balances the lateral pressure
gradient, $\mathcal{O}(10^5~\mbox{cm})$~\citep{spitkovsky2002}.  For
the temporal scales, capturing the rise, $\mathcal{O}(1~\mbox{s})$,
and the decay of the burst, $\mathcal{O}(10~\mbox{s})$, as well as the
accretion period between bursts, $\mathcal{O}(10^4~\mbox{s})$, is
currently beyond the ability of multidimensional hydro codes.
Nevertheless, significant progress has been made in understanding the
multidimensional nature of XRBs, through various approximations.

Laterally propagating detonations were modeled by
\citet{fryxellwoosley82,hedet}. However, since it is difficult to detonate helium at the
densities found in normal XRBs, and even harder to detonate hydrogen
because of the waiting times for weak reactions, these would only occur at
very high densities.  This means that
detonations may only really be applicable to superbursts (where carbon
is the reactant)~\citep{Weinberg2006b,Weinberg2007}.

Global multidimensional studies were performed by
\citet{spitkovsky2002}, where it was demonstrated that the Coriolis
force plays an important role in confining the burning as it spreads
across the neutron star surface.  These calculations used the
shallow-water approximation, so the vertical details of the
atmosphere's structure we not captured.  Their model showed that the
horizontal pressure gradient between the ash and fuel can be important
in accelerating the burning front.

Small-domain studies of convective burning in XRBs preceding flame
development have been done in two-dimensions \citep{lin:2006,xrb,xrb2}
and three-dimensions \citep{xrb3d}.  These calculations used low Mach
number methods, which approximate the hydrodynamics equations to filter
soundwaves, enabling large timesteps and efficient modeling of
subsonic convection.  While these calculations could not support
the lateral differences needed for flame spreading, they can help
understand the role that convection plays in distributing the initial
burning products vertically throughout the neutron star atmosphere as well
as the nature of any turbulence the burning front might encounter as it propagates through the atmosphere.

@@

For deflagrations, we either need to resolve the structure of the
reaction zone or use a flame model.  Flame models usually assume that
the flame structure is thin compared to the size of the system (see,
e.g., \citet{Ropke2007} for applications to Type Ia supernovae).  For
XRBs however, the flame thickness is comparable to the scale height of
the atmosphere, so we cannot use these approximations.  Accurate
models of flames in XRBs therefore require that we resolve the thermal width,
which is $\mathcal{O}(10~\mbox{cm})$ for helium flames~\citep{Timmes00}.

The goal of this study is to understand what numerical and physical
approximations are required to perform a full hydrodynamical,
multidimensional simulation of flame propagation through the
atmosphere of a neutron star.  For simplicity in this first set of
calculations, we will use a pure helium composition.  These studies
complement the prior multidimensional studies described above in helping
us to build a picture of the dynamics of X-ray bursts.

@@

All simulations are performed with the \castro\ hydrodynamics
code~(\citealt{castro}; see also \citealt{astronum:2017} for a recent
description).  We evolve the system of fully compressible Euler
equations for reacting flow:
\begin{eqnarray}
\frac{\partial( \rho X_k)}{\partial t} &=& - \nabla\cdot (\rho \Ub X_k) + \rho \omegadot_k \\
%
\frac{\partial (\rho \Ub)}{\partial t} &=& - \nabla\cdot (\rho \Ub \Ub) - \nabla p +
    \rho \gb \nonumber \\
  &&-2 \rho \Omegab\times \Ub - \Omegab \times (\Omegab \times \rb) \\
%
\frac{\partial (\rho E)}{\partial t} &=& - \nabla \cdot (\rho \Ub E + p\Ub) +
    \nabla \cdot \kth \nabla T + \rho \epsdot +\nonumber\\
  && \rho \Ub \cdot \gb - \rho (\Omegab \cdot \rb)(\Omegab \cdot \Ub) + \rho |\Omegab|^2 (\Ub \cdot \rb)
\end{eqnarray}
Here, $\rho$ is the mass density, $\Ub$ is the velocity, $p$ is the pressure and $E$ is
the specific total energy, which is related to the specific internal
energy as $e = E - |\Ub|^2/2$.  Species are described by mass
fractions, $X_k$ (such that $\sum_k X_k = 1$), and creation rates,
$\omegadot$, and are related to the total specific energy generation
rate, $\epsdot$.  The total mass conservation,
\begin{equation}
\frac{\partial \rho}{\partial t} = - \nabla\cdot (\rho \Ub)
\end{equation}
implies $\sum_k \omegadot_k = 0$.  \castro\ uses an unsplit piecewise
parabolic method (PPM) with characteristic tracing for solving the
hydrodynamics~\citep{ppm,millercolella:2002}, generalized to an
arbitrary equation of state~\citep{zingalekatz}.  Reactions are
incorporated via Strang splitting~\citep{strang:1968}, giving a method
that is overall second-order accurate in space and time.
\castro\ uses the \amrex\ adaptive mesh refinement
library~\citep{amrex_joss} to manage a hierarchy of grids at different
resolutions.

An equation of state of the form $p = p(\rho, e, X_k)$ completes the
thermodynamic description of the system.  Finally, $\kth$ is the
thermal conductivity and $T$ is the temperature.  Diffusion is treated
explicitly in time. \MarginPar{Have not defined what r is in eqs 1-3, and Omega and g are only defined with respect to eq. 5}

@@

We use a general stellar equation of state with nuclei (treated as an
ideal gas), photons, and degenerate/relativistic electrons, as
described in \cite{timmes_swesty:2000}.  To model reactions, we use a 
13-isotope alpha chain, derived from the {\tt aprox13}
network~\citep{timmes_aprox13}.  For one of our runs, we use the
smaller 7-isotope network described in \citet{iso7}.  We integrate the
network using the VODE integration package~\citep{vode}, and our
implementation is provided in the StarKiller Microphysics
source~\citep{starkiller}.
We note that we do not explicitly model viscosity.  The reactions will
provide the smallscale cutoff to the instabilities and turbulence at
the flame front.  We also do not include species
diffusion---astrophysical flames tend to have large Lewis numbers, so
this is not expected to be important~\citep{timmeswoosley:1992}.
Finally, we use the thermal conductivities described in
\citet{Timmes00}.

All simulations use adaptive mesh refinement to refine on the
atmosphere (leaving the space between the top of the atmosphere and
upper boundary at low resolution).  In addition to the base grid, we
use up to 3 refinement levels, the first one a factor of 4 finer than
the previous and the remaining each a factor of 2 finer than the
previous.  \castro\ subcycles in time, so the finer grids are evolved
with a finer timestep than the coarse grids.  Occasionally, the
timestep chosen at the start of a cycle will violate the CFL condition
during the advancement of the finer grids.  In this case, we restart the
finer grid evolution with a smaller timestep, subcycling within the
larger timestep hierarchy.  We use a CFL number of 0.8 for our
simulations.  
% \MarginPar{true for all?} 
The base grid for our standard
simulations is $768\times 192$ zones, and the equivalent finest grid
when 3 refinement levels are added would be $12288\times 3072$ zones.
Our standard domain is $1.2288\times 10^5~\mathrm{cm} \times
3.072\times 10^4~\mathrm{cm}$, corresponding to $10~\mathrm{cm}$ resolution
on the finest grid.  We only refine the fuel layer in the left half of
the domain at the highest resolution (and only down to densities of
$2.5\times 10^4~\gcc$), since this is where we expect the flame to
propagate.  At the start of the simulation, $3.4\%$ of the domain is
at the finest resolution.  This increases to $7.4\%$ by the end of the
simulation, because of the increase in the scale height of the atmosphere
behind the flame.  \MarginPar{add a pic of the grid to show refinement}

Thermal diffusion is modeled explicitly, using a predictor-corrector
scheme to achieve second-order accuracy.  A verification test of the
diffusion scheme is shown in Appendix~\ref{app:diffusion}.  The
explicit thermal diffusion requires a timestep limiter of the form:
\begin{equation}
\Delta t_\mathrm{diff} \le \frac{1}{2} \frac{\Delta x^2}{\mathcal{D}}
\end{equation}
where $\mathcal{D} = \kth/(\rho c_v)$ is the thermal diffusivity.  The
diffusivity increases rapidly at the top of the atmosphere, causing
these low density regions to determine the overall timestep for the
simulations.  Therefore, we disable thermal conduction at low
densities where it is not expected to be important.

We use hydrostatic boundary conditions on the lower boundary, using
a discretized hydrostatic equilibrium equation of the form:
\begin{equation}
\label{eq:hse}
p_i = p_{i-1} + \frac{1}{2} \Delta y (\rho_i + \rho_{i-1}) g
\end{equation}
and holding the temperature constant in the ghost cells.  This is solved
together with the equation of state.  The velocity is reflected at this boundary.
This procedure follows the form described in \citet{ppm-hse}.  The
left boundary is reflecting and the right boundary is a zero-gradient
outflow.  The top boundary sets the state to simply the conditions
in our outer buffer region of the initial model (see below), with the
normal velocity set to the larger of zero or the velocity at the top
of the domain (this prevents incoming velocities at the top) and the
transverse velocities set to zero.

When we begin the simulation, there is a transient phase as the flame
gets established.  Material that is forced upward will encounter the
steep density gradient at the top of the atmosphere and accelerate as
it is blown out of the atmosphere.  Eventually this material will fall
back to the top of the atmosphere.  In this paper, we are mostly
concerned with the behavior of the flame and not any material that is
violently blown out of the top of the atmosphere, so we apply a sponge
to this region.  This is similar to the method we previously used in \cite{xrb3d}.
The sponge drives the velocity of the material in the low
density \MarginPar{add mathematical form of sponge} regions at the top
of the atmosphere to zero.  This sponging helps increase our timestep
as well.

@@

The speed and thickness of a laminar helium flame are determined by the
energy generation rate and conductivity, and scale roughly as 
\begin{equation}
v_f \approx \sqrt{k_\mathrm{th} \dot{\epsilon}} \qquad
\lambda_f \approx \sqrt{\frac{k_\mathrm{th}}{\dot{\epsilon}}}
\end{equation}
\citep{orourke:1979,khokhlov:1993}.
At the densities we consider in this simulation, the pure He flame
speed is quite slow and would require a prohibitively long integration
time to see significant evolution of the burning.  We therefore
consider boosted flames in this first paper, to accelerate the burning
and allow us to understand the qualitative effects of laterally
propagating flames.  To boost the flame while keeping the thickness
the same, we can multiply both the burning rate and the conductivity
by the same factor.  For our standard calculations, we choose 10 for
each, to give a 10$\times$ faster flame speed.  We call this the ``10/10''
flame.  We will also do a simulation with the reactions and
conductivity both boosted by 5, the ``5/5'' flame.

@@

The species mass fractions use this same profile, switching from a set
describing the underlying star, ${X_k}_\star$, and the set for the
accreted material, ${X_k}_\mathrm{atm}$, which is used in the
isentropic and outer regions.  Note, since the profile above is
linear in $X$, if the initial mass fractions sum to one, then the blended
mass fractions in the ramp region also sum to one.

@@

Creating the model involves specifying $T_\star$, $T_\mathrm{hi}$,
$T_\mathrm{lo}$, $\rho_\mathrm{int}$, $H_\star$,
$\delta_\mathrm{atm}$, ${X_k}_\star$, and ${X_k}_\mathrm{atm}$.  We
then integrate outwards from the base of the ramp region ($y =
H_\star$), enforcing the discrete form of hydrostatic equilibrium,
Eq.~\ref{eq:hse}.  Integrating upwards, we find $p_i$ and
$\rho_i$ using a Newton solve together with the equation of state, with either the temperature
specified, $T_i = T(p_i, \rho_i, \{X_k\}_i)$ (in the isothermal, ramp,
and buffer layers) or constant entropy, $s_i = s(p_i, \rho_i,
\{X_k\}_i)$, in the fuel layer.  This follows the procedures described in \citet{ppm-hse}.  We
use a constant temperature for all $y < H_\star +
3\delta_\mathrm{atm}$.  Above this, we switch to an isentropic atmosphere until the
temperature drops to a floor value, $T_\mathrm{lo}$, at which point we
again keep the temperature constant.  The integration of the
atmosphere continues until the density falls to a low density cutoff,
$\rho_\mathrm{cutoff}$.  The material above this height is taken to
have constant density and temperature.

@@

We create two models, a ``cool'' model representing the atmosphere
ahead of the flame and a ``hot'' (or perturbed) model that will
initiate the flame by driving the burning.  \MarginPar{This has already been said at the start of this section} The parameters we use for
the model generation are listed in Table~\ref{table:params}. \MarginPar{reference and discuss the figure of initial model here}

@@

Table~\ref{table:sim_names} summarizes the simulations that we ran. \MarginPar{This table has different model names to the previous table}
The majority of them used a reaction rate boosting of $10$ and a
conductivity boosting of $10$, which should increase the flame speed
by a factor of $10$.  Most simulations used a resolution of
$10~\mbox{cm}$ and a domain width of a little more than one kilometer.  
We use an artifically high rotation rate of
$2000~\mbox{Hz}$, which gives a Rossby length of
\begin{equation}
L_R = \frac{\sqrt{g H_0}}{\Omega} \sim 3\times 10^4~\mathrm{cm}
\end{equation}
using a scale height $H_0 = 10^3~\mathrm{cm}$.  This is about one
quarter of the domain width.  The run at $20~\mbox{cm}$ resolution used
one fewer level of refinement.  The slower rotating case using a
slightly wider domain to accommodate the expected larger Rossby
length.  We note that the entire simulation framework for these
calculations is freely available in the \castro\ github
repository\footnote{\url{https://github.com/AMReX-Astro/Castro}, using
  the {\tt flame\_wave} setup.}.  In the discussions below, we'll use
the simulation names defined in the table to refer to specific runs
and we'll use the 10/10 run as the reference calculation.

@@

Figure~\ref{fig:time_series} shows the time evolution of the 10/10
simulation, focusing on the mean-molecular weight,
\begin{equation}
\bar{A} = \left ( \sum_k \frac{X_k}{A_k} \right )^{-1}.
\end{equation}
In each frame the buffer of \isot{Ni}{56} that serves as the
underlying neutron star is seen spanning the bottom of the domain.
Above that, the composition begins as pure \isot{He}{4}, but as the
simulation progresses, the flame processes this to heavier nuclei,
increasing $\bar{A}$.  By about $10~\mbox{ms}$, we see the flame front
is reasonable well-defined.  We see that the bottom of the burning
front is lifted off of the base of the atmosphere, greatly increasing
the surface area of the burning compared with a perfectly vertical
flame front.  By $20~\mbox{ms}$ the flame has moved out substantially
and we are beginning to see a gradient in the composition of the ash,
with the heavier nuclei furthest behind the flame.  The boosting of
the burning likely artificially increases this effect. \MarginPar{Doesn't the increased flame speed of the 10/10 run vs. the 5/5 run show this?}

@@

A final feature worth noting is the ash that seems to move along the
surface at a higher velocity than the flame, via surface gravity
waves.  The sponging that we perform is likely damping this to some extent, and
the method by which we initialize the flame may induce a larger
transient than in nature. Nevertheless, this surface ash is intriguing
because it affects the composition of the photosphere ahead of the
burning front, potentially changing our interpretation of
observations.  This is something that will be explored more fully in
the future.

@@

Our default resolution puts $\sim 80$ zones vertically in the
``cool'' \MarginPar{double check} model atmosphere height.  To
understand the effects of resolution, we also performed a run with one
fewer refinement level, giving a $20~\mbox{cm}$ resolution overall (this is our
10/10-lores run).  Figure~\ref{fig:10_10_lowres} shows the fields for this run at
20~ms.  The structure is largely the same as the 10/10 run, with
largely the same flame shape and position, and the same structure in
the energy generation rate.  Since we do not have a jump in refinement
right below the atmosphere, we don't see the temperature feature from
the initial model mapping there, but we do see some cooling in the
\isot{Ni}{56} region.  This is likely a resolution effect, again due
to the strong gradient in temperature at the base of the atmosphere.
The strong agreement in the low resolution run to the 10/10 run gives
us confidence that we are capturing the flame physics properly.

@@

We also considered the effect of the network size.
Figure~\ref{fig:network} shows a comparison of the 10/10 boosting run
with the standard 13-isotope ``aprox13'' network and the reduced
7-isotope ``iso7'' network.  We see that the flame in the aprox13 case
is slightly more advanced and has a higher $\bar{A}$ than the iso7
case.  We see in the next section that these two networks give largely
the same flame speed.  The reduced network size saves a lot of memory,
which will be useful when we transition to 3-d simulations.
